Learning words by recursive pragmatic reasoning about other agents

### 2013-05-31 praglearn.tex
simulation text edits

Experimental results suggest that communicators are able to establish novel, consensus-based communication systems. For example, adults playing a communication game using only novel symbols with no conventional meaning will typically converge on a set of new conventions which allow them to accomplish their task \cite{galantucci2005}. In a second experiment, communicators asked to refer to novel objects through the course of the interaction invent conventional names for them (e.g., ``the ice skater'' for an abstract figure vaguely resembling an ice skater, \cite{clark1986}). From a pure learning perspective this is anomalous: since both agents know perfectly well that there is no underlying convention to discover, there is nothing to learn from the other's behaviour. Furthermore, even if only one partner is producing the novel expressions, their behaviour still becomes more regular (conventional) over time, which would seem to rule out a role for learning -- even if there is some pattern in the expressions the speaker chooses to use, there is certainly nothing for the \textit{speaker} to learn by observing these patterns, and thus their behaviour should not change over time. An obvious hack might be to simply stipulate that speakers take their own utterances as training data for further learning, but this would fail to explain yet another wrinkle in this phenomenon: speakers only regularize their production over time if they have some kind of feedback indicating that the listener understands; if such feedback is removed, the phenomenon is dramatically reduced (XX krauss 1966, branigan catchpole pickering 2011).

@@

\begin{SCfigure}
\centering
\includegraphics[width=0.3\textwidth]{figures/emergence-average.pdf}
\hspace{0.2in}
\includegraphics[width=0.3\textwidth]{figures/emergence-horn-average.pdf}
\caption{\label{fig:emergence-average} Averaged behaviour over 300 dialogues as in Figs.~\ref{fig:emergence} and \ref{fig:horn}. Left: Communicative success by game type and dialogue turn. Right: Proportion of dyads in the Horn implicature game (\S\ref{sec:horn-emergence}) who have converged on the ``good'' or ``bad'' lexicons and believe that these are literal meanings.}
\end{SCfigure}

To model such phenomena, we imagine two agents playing the simple referential game introduced above, in which on each turn the speaker is assigned a target object, they utter some word referring to this object, the listener makes a guess at the object, and then, critically, the speaker observes the listener's guess and the listener receives feedback indicating the correct answer (i.e., the speaker's intended meaning), which they use to update their posterior over lexicons before proceeding to the next trial. As in \cite{krauss1964,clark1986}, the speaker and listener remain fixed in the same role throughout. Fig.~\ref{fig:emergence} shows the result of simulating several such games when both parties begin in total ignorance. Notice that: (a) agents' performance begins at chance, but quickly rises; (b) they tend towards structured, sparse lexicons with a one-to-one correspondence between objects and words; (c) as the speaker and listener have entirely different data (the listener's interpretations and the speaker's intended meaning, respectively), unlucky early guesses can lead them to believe in entirely contradictory lexicons -- but they generally recover and converge despite this.

What's happening here is that each agent is, effectively, using their partner's behaviour as a basis for forming superstitions about the underlying lexicon that they assume must exist. Since they then each act on these superstitions, and their partner uses the resulting actions to form new superstitions, they soon converge on using similar lexicons, and what started as a superstition becomes normatively correct.

What happens if we remove feedback from the listener to the speaker? If the speaker is not allowed to observe the listener's guess, then the speaker does no learning and their posterior over lexicons remains perfectly uniform. They therefore choose words by simply flipping a coin, and performance remains at chance.

The role of learning here is obvious, but what does the pragmatics part of the model do? First, it makes it possible for the speaker to learn from listener interpretations, which is not possible without an explicit model of the speaker as a rational agent. Second, it creates the sparse structure in these observed lexicons: at each step, the goal-directed speaker attempts to exploit whatever communicatively useful asymmetries exist in its current model of the listener, and these attempts tend to be successful, since if the listener has even a small asymmetry in their current posterior then their pragmatic reasoning will magnify it as well. But over time, the speaker's use of such patterns, and the listener's ability to understand them, will each be taken by the other as evidence that these patterns actually exist in the literal lexicon. The result is that while both agents start out with uniform priors over the lexicon, they are biased to end up believing in lexicons that are communicatively useful for the task at hand.

@@

\begin{SCfigure}
\centering
\includegraphics[width=0.5\textwidth]{figures/horn-composite.pdf}
% \includegraphics[width=0.48\textwidth]{figures/horn-emergence-1.pdf}
\caption{\label{fig:horn} Dialogue.}
\end{SCfigure}

A stronger example of how pragmatics can create such biases can be observed if we simulate agents playing this naming game in the ``cheap''/``expensive'' \textsc{common}/\textsc{rare} domain of \S\ref{sec:horn-implicature}. Recall that in this domain, pragmatic reasoning causes each agent to start out independently biased -- but only weakly -- towards the associations ``cheap'' $\leftrightarrow$ \textsc{common}, ``expensive'' $\leftrightarrow$ \textsc{rare}, and that this bias arises exactly because their prior on literal meaning is uniform and uninformative. A fully rational listener, therefore, observing our uncertain speaker using this mapping, would discount it as arising from this bias, and conclude that the speaker in fact did not know which convention was in use. Our socially anxious listener, however, believes that the speaker does know which convention is in use, and therefore tends to take their behaviour as evidence that the good system is in fact in use. Similarly, the socially anxious speaker will wager that since on average they will succed more often if the listener is using the good system, they might as well try it -- and when they succeed, they take this as evidence that the listener was in fact using the good system all along. As a result, dyads in this game not only end up converging onto a stable system, but onto specifically the good system, at a rate far above chance. However, something fundamentally different happens here than in the specificity implicature case (\S\ref{sec:learning-specificity-implic}). There, we saw that our model could learn hear pragmatically strengthened usages of words while recovering the underlying literal meaning. In this case, though, the pragmatic strengthening depends essentially on ignorance about literal meaning. As the agents gather more data, therefore, the implicature disappears, to be replaced by a belief that ``cheap'' \textit{literally} means \textsc{common}. To demonstrate this, we queried each agent in each dyad about how they would refer to or interpret each object and word, \textit{if} the two objects were equally common, which cancels the Horn implicature. As shown in Fig.~\ref{emergence-average} (right), after 30 turns, in nearly 70\% of dyads both $S$ and $L$ believed in the ``good'' mapping even in this implicature-free case, while less than 20\% believed in the ``bad'' mapping (with the rest being inconsistent).

>>>>

Experimental results suggest that communicators are able to establish novel, consensus-based communication systems. For example, adults playing a communication game using only novel symbols with no conventional meaning will typically converge on a set of new conventions which allow them to accomplish their task \cite{galantucci2005}. Or in a less extreme example, communicators asked to refer to novel objects invent conventional names for them over the course of repeated interactions (e.g., ``the ice skater'' for an abstract figure vaguely resembling an ice skater, \cite{clark1986}). From a pure learning perspective this is anomalous, however: Since both agents know perfectly well that there is no underlying convention to discover, there is nothing to learn from the other's behaviour. Furthermore, even if only one partner is producing the novel expressions, their behaviour still becomes more regular (conventional) over time, which would seem to rule out a role for learning---even if there is some pattern in the expressions the speaker chooses to use, there is certainly nothing for the \textit{speaker} to learn by observing these patterns, and thus their behaviour should not change over time. 
% An obvious hack might be to simply stipulate that speakers take their own utterances as training data for further learning, but this would fail to explain yet another wrinkle in this phenomenon: speakers only regularize their production over time if they have some kind of feedback indicating that the listener understands; if such feedback is removed, the phenomenon is dramatically reduced (XX krauss 1966, branigan catchpole pickering 2011).

@@

To model such phenomena, we imagine two agents playing the simple referential game introduced above. On each turn the speaker is assigned a target object, utters some word referring to this object, the listener makes a guess at the object, and then, critically, the speaker observes the listener's guess and the listener receives feedback indicating the correct answer (i.e., the speaker's intended meaning). The listener then updates their posterior over lexicons before proceeding to the next trial. As in \cite{krauss1964,clark1986}, the speaker and listener remain fixed in the same role throughout.

Fig.~\ref{fig:emergence} shows the result of simulating several such games when both parties begin in total ignorance. Notice that: (a) agents' performance begins at chance, but quickly rises; (b) they tend towards structured, sparse lexicons with a one-to-one correspondence between objects and words; and (c) as the speaker and listener have entirely different data (the listener's interpretations and the speaker's intended meaning, respectively), unlucky early guesses can lead them to believe in entirely contradictory lexicons---but they generally recover and converge. Each agent effectively uses their partner's behavior as a basis for forming weak beliefs about the underlying lexicon that they assume must exist. Since they then each act on these beliefs, and their partner uses the resulting actions to form new superstitions, they soon converge on using similar lexicons, and what started as a ``superstition'' becomes normatively correct. 
%On the other hand, if the speaker is not allowed to observe the listener's guess, then the speaker does no learning and their posterior over lexicons remains perfectly uniform. They therefore choose words by simply flipping a coin, and performance remains at chance.

Pragmatic inference here does two things. First, the pragmatic recursion makes it possible for the speaker to learn from the listener's interpretations, which would be impossible without an explicit model of the listener as a rational agent. Second, pragmatics creates the sparse structure in the observed lexicons: At each step, the goal-directed speaker attempts to exploit whatever communicatively useful asymmetries exist in its current model of the listener, and these attempts tend to be successful. If the listener has even a small asymmetry in their current posterior then their pragmatic reasoning will magnify it as well.

% But over time, the speaker's use of such patterns, and the listener's ability to understand them, will each be taken by the other as evidence that these patterns actually exist in the literal lexicon. The result is that while both agents start out with uniform priors over the lexicon, they are biased to end up believing in lexicons that are communicatively useful for the task at hand.

@@

\begin{SCfigure}
\centering
\includegraphics[width=0.5\textwidth]{figures/horn-composite.pdf}
% \includegraphics[width=0.48\textwidth]{figures/horn-emergence-1.pdf}
\caption{\label{fig:horn} Example simulations showing the lexicalization of horn implicatures. Plotting conventions are as above.}
\end{SCfigure}

\begin{SCfigure}
\centering
\includegraphics[width=0.3\textwidth]{figures/emergence-average.pdf}
\hspace{0.2in}
\includegraphics[width=0.3\textwidth]{figures/emergence-horn-average.pdf}
\caption{\label{fig:emergence-average} Averaged behaviour over 300 dialogues as in Figs.~\ref{fig:emergence} and \ref{fig:horn}. Left: Communicative success by game type and dialogue turn. Right: Proportion of dyads in the Horn implicature game (\S\ref{sec:horn-emergence}) who have converged on the ``good'' or ``bad'' lexicons and believe that these are literal meanings.}
\end{SCfigure}
A stronger example of how pragmatics can create such biases can be observed if we simulate agents playing this naming game in the ``cheap''/``expensive'' \textsc{common}/\textsc{rare} domain of \S\ref{sec:horn-implicature}. Recall that in this domain, pragmatic reasoning causes each agent to start out independently biased -- but only weakly -- towards the associations ``cheap'' $\leftrightarrow$ \textsc{common}, ``expensive'' $\leftrightarrow$ \textsc{rare}, and that this bias arises exactly because their prior on literal meaning is uniform and uninformative. A fully rational listener, therefore, observing our uncertain speaker using this mapping, would discount it as arising from this bias, and conclude that the speaker in fact did not know which convention was in use. Our socially anxious listener, however, believes that the speaker does know which convention is in use, and therefore tends to take their behaviour as evidence that the good system is in fact in use. Similarly, the socially anxious speaker will wager that since on average they will succed more often if the listener is using the good system, they might as well try it -- and when they succeed, they take this as evidence that the listener was in fact using the good system all along. As a result, dyads in this game not only end up converging onto a stable system, but onto specifically the good system, at a rate far above chance. However, something fundamentally different happens here than in the specificity implicature case (\S\ref{sec:learning-specificity-implic}). There, we saw that our model could learn hear pragmatically strengthened usages of words while recovering the underlying literal meaning. In this case, though, the pragmatic strengthening depends essentially on ignorance about literal meaning. As the agents gather more data, therefore, the implicature disappears, to be replaced by a belief that ``cheap'' \textit{literally} means \textsc{common}. To demonstrate this, we queried each agent in each dyad about how they would refer to or interpret each object and word, \textit{if} the two objects were equally common, which cancels the Horn implicature. As shown in Fig.~\ref{emergence-average} (right), after 30 turns, in nearly 70\% of dyads both $S$ and $L$ believed in the ``good'' mapping even in this implicature-free case, while less than 20\% believed in the ``bad'' mapping (with the rest being inconsistent).
